% Auto-extracted body of preregistration/preregistration.tex
% for inclusion in the main paper appendix via \input{}.
\begin{center}
\begin{tabular}{@{}rl@{}}
  \textbf{Date of pre-registration:} & [DATE] \\
  \textbf{Design matrix SHA-256:}    & \texttt{31a0dc197b46...df1728d6} \\
  \textbf{Software:}                 & Python 3.11, NumPy 2.3.5, SciPy 1.16.3, seed\,$=$\,42 \\
\end{tabular}
\end{center}

% ================================================================
%  Abstract
% ================================================================

\paragraph{Summary of the pre-registration.}
We pre-register a controlled factorial evaluation of Retrieval-Augmented
Generation (RAG) pipeline composition.  The study commits to three
confirmatory hypotheses, a 242-configuration D-optimal fractional
factorial design spanning 8 pipeline dimensions, 5 benchmark datasets
(2{,}300 queries per configuration), and a three-track evaluation
protocol covering retrieval, end-to-end generation, and hallucination
resistance.  To our knowledge, this is the first pre-registered
factorial evaluation of RAG pipeline composition---$4.6\times$ the
scale of the largest prior controlled RAG
study~\citep{jin2024flashrag}.  We publish this
document before experimental data collection to establish hypothesis
commitment and invite community feedback on model selection, benchmark
composition, and statistical methodology.
% ================================================================
\subsection{Motivation and Rationale}
% ================================================================

Retrieval-Augmented Generation~\citep{lewis2020rag} has become the
dominant paradigm for grounding large language models in external
knowledge.  A rich toolkit of individually validated
techniques---dense retrieval, hybrid sparse--dense fusion, cross-encoder
reranking, HyDE~\citep{gao2023hyde}, CRAG~\citep{yan2024crag},
Self-RAG~\citep{asai2023selfrag}---is available to practitioners, yet
each technique is evaluated in isolation against a simple baseline.
Practitioners who combine several techniques simultaneously have no
empirical basis for predicting whether their gains add, cancel, or
amplify.

This is not merely a reporting gap.
\citet{wang2024searching} observe that end-to-end RAG performance
depends on coordinated pipeline design rather than any single component;
\citet{chen2025mmoarag} confirm the presence of interaction effects by
showing that jointly optimized RAG components outperform independently
tuned ones.  Yet neither work measures the sign, magnitude, or
domain-dependence of these interactions under controlled factorial
conditions---precisely the question a practitioner faces when choosing a
configuration for production.

Pre-registration prevents post-hoc hypothesis mining, a known problem in
ML benchmarking~\citep{dodge2019importance,bouthillier2021accounting}.
By committing to hypotheses, analysis plans, and exclusion criteria
before data collection, we ensure that our confirmatory findings are
distinguishable from exploratory ones.  We additionally publish this
document to invite community input on experimental choices that remain
open (Section~\ref{prereg:sec:models}).

% ================================================================
\subsection{Pre-Registered Hypotheses}
\label{prereg:sec:hypotheses}
% ================================================================

\paragraph{H1 --- Interaction Dominance.}
Two-way interactions between the 8 factorial dimensions explain
$\geq$20\% of the total variance in faithfulness (RAGAS, 0--1 scale)
across the factorial sweep, after accounting for main effects.
Variance is measured via partial~$\eta^2$ in a Type~III ANOVA.
\textit{Test:} F-test for interaction terms, $\alpha = 0.05$.
\textit{Disconfirming result:} If interaction terms contribute
$<$5\% of variance, RAG components are approximately additive and
can be optimized independently.

\paragraph{H2 --- Pareto Dominance of Balanced Profile.}
The \textsc{balanced} operational profile is Pareto-dominant over the
\textsc{max\_quality} profile in $\geq$3 of the 5 benchmark datasets,
where dominance is defined as strictly higher hypervolume indicator
(HV) in the 3-objective space (faithfulness, latency, cost).
\textit{Test:} Wilcoxon signed-rank on per-query faithfulness scores,
Bonferroni-corrected ($\alpha = 0.05 / 5 = 0.01$ per dataset).
\textit{Disconfirming result:} If \textsc{max\_quality} dominates in
$\geq$4/5 datasets, feature saturation and interaction penalties are
weaker than expected.

\paragraph{H3 --- Corpus Scaling Saturation.}
For dense-only retrieval, NDCG@10 exhibits power-law saturation:
a fitted curve $\text{NDCG@10} = a \cdot N^b + c$ yields
$b < 0.05$ for $N \geq 100$K documents, analogous to the saturation
behavior observed in neural scaling laws~\citep{kaplan2020scaling}.  For hybrid retrieval, the
saturation threshold $N^*$ is strictly lower than for dense-only
retrieval on at least 3/5 benchmarks.
\textit{Test:} Bootstrap 95\% CI on $b$ (10{,}000 resamples);
permutation test (10{,}000 permutations) comparing $N^*$.
\textit{Disconfirming result:} If dense retrieval does not saturate
($b \geq 0.10$), embedding models scale better than expected with
corpus size.

% ================================================================
\subsection{Experimental Design}
\label{prereg:sec:design}
% ================================================================

% -- Platform --------------------------------------------------
\subsubsection{Platform}

\textsc{Lasagna} is a stage-isolated RAG evaluation platform: each
pipeline stage is implemented behind an abstract base class, and
configuration changes to one stage do not affect the execution of any
other stage.  This architectural invariant guarantees that the output
of stage~$i$ is a deterministic function of the output of stage~$i{-}1$
and the configuration parameters~$\boldsymbol{\theta}_i$ alone,
eliminating the coupling confounds that invalidate factorial
interpretation in conventional RAG implementations.

% -- Candidate Models ------------------------------------------
\subsubsection{Candidate Models}
\label{prereg:sec:models}

All configurations share a single generator LLM (held constant) and a
single evaluation judge.  We list primary and alternative candidates
to invite community input before experiments begin.

\begin{table}[h]
\centering
\footnotesize
\caption{Model candidates.  The generator is held constant across all
  configurations; the judge evaluates all RAGAS metrics.  Embedding
  models are factorial dimension~D1.}
\label{prereg:tab:models}
\begin{tabular}{@{}p{2.0cm}p{3.4cm}p{3.8cm}p{4.0cm}@{}}
\toprule
\textbf{Role} & \textbf{Primary} & \textbf{Alternatives} & \textbf{Selection rationale} \\
\midrule
Generator    & GPT-4.1 (OpenAI)
             & Claude~3.7 Sonnet (Anthropic), Gemini~2.0 (Google)
             & Held constant; cost vs.\ capability \\
\addlinespace
Judge        & GPT-4o-mini (OpenAI)
             & Claude~3.5 Haiku (Anthropic), Llama-3-70B (Meta)
             & Determinism at $T{=}0$; cost for ${\sim}$1.7M calls \\
\addlinespace
Embedding    & \multicolumn{2}{l}{MiniLM-L6 (384d), BGE-M3 (1024d), E5-large (1024d), BGE-large (1024d)}
             & Span lightweight to SOTA multilingual \\
\addlinespace
Reranker     & ms-marco-MiniLM-L6-v2
             & BGE-reranker-v2
             & Standard BEIR cross-encoder baseline \\
\bottomrule
\end{tabular}
\end{table}

We welcome suggestions on model selection before experiments begin.
In particular, we are open to: (1)~alternative generator models that
would increase reproducibility (e.g., open-weight models),
(2)~additional judge models for cross-validation of faithfulness
scores, and (3)~embedding models that would expand the capacity range.

% -- Factorial Dimensions --------------------------------------
\subsubsection{Factorial Dimensions}

\begin{table}[h]
\centering
\footnotesize
\caption{The eight factorial dimensions.  Baseline levels (marked~$^*$)
  are the simplest, lowest-cost options.}
\label{prereg:tab:dimensions}
\begin{tabular}{@{}llcl@{}}
\toprule
Dim & Stage & $k$ & Levels \\
\midrule
D1 & Embedding      & 4 & MiniLM-L6$^*$, BGE-M3, E5-large, BGE-large \\
D2 & Chunking       & 4 & fixed$^*$, semantic, sentence-window, hierarchical \\
D3 & Enrichment     & 4 & none$^*$, contextual-retrieval, RAPTOR, GraphRAG \\
D4 & Retrieval      & 3 & dense$^*$, hybrid-BM25, hybrid-SPLADE \\
D5 & Query aug.     & 5 & none$^*$, HyDE, RAG-Fusion, decomposition, step-back \\
D6 & Reranking      & 4 & none$^*$, single-CE, ensemble, MMR \\
D7 & Compression    & 2 & none$^*$, LLMLingua \\
D8 & Agentic gate   & 4 & none$^*$, CRAG, Self-RAG, Speculative-RAG \\
\bottomrule
\end{tabular}
\end{table}

The full factorial comprises
$4 \times 4 \times 4 \times 3 \times 5 \times 4 \times 2 \times 4 =
30{,}720$ configurations.

% -- D-Optimal Design -----------------------------------------
\subsubsection{D-Optimal Design}

We reduce 30{,}720 candidates to \textbf{242 configurations} using a
D-optimal fractional factorial generated by Fedorov coordinate
exchange~\citep{fedorov1972theory,atkinson2007optimum}.  The model
matrix includes 1~intercept $+$ 22~main-effect $+$ 209~two-way
interaction columns $=$ 232 parameters, with 10 residual degrees of
freedom.  Effect coding ensures orthogonal main
effects~\citep{box2005statistics,montgomery2017design}.  All 232 model
terms are estimable (rank verification), and no two-way interaction
correlates with any representative three-way interaction at
$|r| > 0.30$.

\textbf{Design matrix SHA-256:}
\texttt{31a0dc197b46a90819fc9178ea06668444ad9617ff4c8a86f3d557a4df1728d6}
(seed\,$=$\,42).

% -- Benchmarks ------------------------------------------------
\subsubsection{Benchmark Datasets}

\begin{table}[h]
\centering
\footnotesize
\caption{Planned benchmark datasets.  Total evaluation queries per
  configuration: 2{,}300.}
\label{prereg:tab:benchmarks}
\begin{tabular}{@{}llrrl@{}}
\toprule
ID & Dataset & Corpus & Eval queries & Relevance \\
\midrule
DS1 & MS~MARCO~v1     & 8.8M passages       & 500 sampled & BM25 sparse (TREC) \\
DS2 & HotpotQA        & 5.2M paragraphs      & 500 sampled & Supporting facts \\
DS3 & BEIR/FiQA       & 57K posts            & 500 sampled & Crowdsourced \\
DS4 & BEIR/SciFact    & 5K abstracts         & 300 (all)   & Expert binary \\
DS5 & CUAD (legal)    & 510 contracts        & 500 sampled & Clause-level \\
\bottomrule
\end{tabular}
\end{table}

These five benchmarks span general web (DS1), multi-hop reasoning
(DS2), financial opinion QA (DS3), scientific claim verification
(DS4), and legal contract analysis (DS5), covering diversity in domain,
relevance granularity, and query
complexity~\citep{thakur2021beir}.

% -- Metrics ---------------------------------------------------
\subsubsection{Metrics}

\textit{Retrieval:} NDCG@10, MRR, Recall@100.
\textit{End-to-end:} faithfulness, answer relevance
(RAGAS~v0.4~\citep{es2023ragas}, judged by GPT-4o-mini at $T{=}0$),
exact match, token~F1, p50/p95 latency, \$/query.
\textit{Multi-objective:} hypervolume indicator~(HV) in the
faithfulness--latency--cost space, reference point
$(0, 10{,}000\,\text{ms}, \$1.00)$.

% ================================================================
\subsection{Statistical Analysis Plan}
\label{prereg:sec:analysis}
% ================================================================

\paragraph{H1.}
Type~III ANOVA with the model:
{\small
\begin{verbatim}
faithfulness ~ D1+D2+D3+D4+D5+D6+D7+D8
             + D1:D2+D1:D3+...+D7:D8   (28 two-way terms)
\end{verbatim}
}
\vspace{-0.5em}
Test statistic: F for the joint set of all interaction terms,
$F(209,\, n_\text{resid})$.  Effect size: partial
$\eta^2 = \text{SS}_\text{int} / (\text{SS}_\text{int} +
\text{SS}_\text{resid})$; threshold $\geq 0.20$.
Post-hoc: Bonferroni ($\alpha = 0.05/28 = 0.00179$) per interaction
pair.

\paragraph{H2.}
Wilcoxon signed-rank test on per-query faithfulness
(\textsc{balanced} vs.\ \textsc{max\_quality}, paired by query).
Bonferroni across 5~datasets ($\alpha = 0.01$ per dataset).
HV computed with exact 3D algorithm; objectives normalized to $[0,1]$.

\paragraph{H3.}
Nonlinear least-squares fit of $\text{NDCG@10} = a \cdot N^b + c$
using \texttt{scipy.optimize.curve\_fit}.  Confirmed if upper 95\%
bootstrap CI on $b < 0.05$ for $N \geq 100$K.  Comparison of $N^*$
via permutation test (10K permutations).

\paragraph{Power.}
For $n = 500$ queries with $\alpha_\text{corrected} = 0.00179$,
minimum detectable effect at 80\% power:
$d \approx 0.168$ (Cohen's $d$).

% ================================================================
\subsection{Exclusion Criteria}
\label{prereg:sec:exclusion}
% ================================================================

\begin{enumerate}[nosep,leftmargin=1.5em]
  \item Queries exceeding 30\,s wall-clock are excluded and counted.
  \item Configurations with $>$10\% null-retrieval queries are marked
    \textsc{Failed} and excluded from the ANOVA.
  \item Corpus sizes where index construction fails (OOM or $>$24\,h)
    are excluded from the scaling curve fit.
  \item Configurations require $\geq$200 valid queries (of 2{,}300)
    to enter the ANOVA.
\end{enumerate}

% ================================================================
\subsection{Budget, Timeline, and Deviations Protocol}
\label{prereg:sec:budget}
% ================================================================

\begin{table}[h]
\centering
\footnotesize
\caption{Estimated compute budget.}
\label{prereg:tab:budget}
\begin{tabular}{@{}lrrr@{}}
\toprule
Experiment & Configs & Queries/config & Est.\ cost (USD) \\
\midrule
Factorial sweep     & 242 & 2{,}300 & \$2{,}478 \\
Scaling experiment  & 15  & 500     & \$30 \\
Pareto experiment   & 20  & 500     & \$45 \\
\midrule
\textbf{Total}      &     &         & $\mathbf{\sim\$2{,}553}$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{Timeline.}
All experiments will be run \textit{after} archival of this document.
Week~1--3: factorial sweep (H1).
Week~4: scaling experiment (H3).
Week~4--5: Pareto experiment (H2).
Week~5--6: analysis and paper writing.

\textbf{Deviations protocol.}
Any deviation from this pre-registration will be: (1)~documented in a
dated amendment, (2)~reported transparently in the paper, and
(3)~accompanied by both the pre-registered analysis and the modified
analysis so that reviewers can assess the impact of the deviation.
Any analysis not specified here will be labeled \textit{exploratory}.

% ================================================================
\subsection*{Community Input}
% ================================================================

This pre-registration is published before experimental data collection
to establish hypothesis commitment and invite community input.
We welcome feedback on:
\begin{enumerate}[nosep,leftmargin=1.5em]
  \item additional hypotheses worth testing within this factorial
    framework,
  \item model selection for generation and evaluation
    (Section~\ref{prereg:sec:models}),
  \item benchmark composition and domain coverage,
  \item statistical methodology and effect-size thresholds.
\end{enumerate}
All community-suggested modifications adopted before experiments begin
will be documented as amendments; post-experiment suggestions will be
addressed as exploratory analyses in the companion paper.

% ================================================================
